Global patterns of forest autotrophic carbon fluxes

### 2020-07-06 manuscript/manuscript draft.Rmd
updated manuscript draft with new references

Carbon cycling in forests worldwide provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity and many human populations (**[MEA 2005](http://www.millenniumassessment.org/documents/document.354.aspx.pdf)**; **REFS**), while strongly influencing atmospheric CO~2~ and climate [@bonan_forests_2008]. Forests' autotrophic carbon fluxes (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (e.g., **[Zak et al. 1994](http://esajournals.onlinelibrary.wiley.com/doi/abs/10.2307/1940888)**; [**Niedziałkowska et al. 2010](https://doi.org/10.2980/17-1-3310)**) and possibly diversity [@waide_relationship_1999; **Chu et al. 2019 (DOI: 10.1111/ele.13175)**]. FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients.

@@

The latitudinal gradient in FACF, along with altitudinal gradients [@girardin_net_2010; @malhi_variation_2017], is primarily driven by climate, which is a significant driver of FACF across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011; @hursh_sensitivity_2017]. The majority of studies have focused on exploring the relationships of FACF to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. There is strong evidence that both MAT and MAP show significant positive relationships with FACF [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of FACF across studies. Various measures of primary productivity {**FACF?**} saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity {**FACF?**} are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while $GPP$ increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] finds that $NPP$ increases linearly with temperature. Taylor et al. [-@taylor_temperature_2017] showed a positive interaction between MAT and MAP in shaping tropical forest productivity, such that high rainfall had a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates. Such complicated dynamics play out with belowground fluxes as well, with typically dominant soil temperature mediated by moisture and carbon supply (**Hursh et al. 2017, http://dx.doi.org/10.1111/gcb.13489; Xu et al. 2016 http://dx.doi.org/10.1016/j.jplph.2016.08.007**).

@@

This analysis focused on mature forests with no known history of signficant disturbance or management. 
There is evidence that stand age influences patterns of FACF and carbon allocation in forest ecosystems, and can confound relationships between latitude and primary productivity [@delucia_forest_2007; @gillman_latitude_2015]. To reduce any biasing effects of stand age, we included only stands of known age $\ge$ 100 years and those described by terms such as "mature", "intact", or "old-growth". 
Since management can alter observed patterns of FACF [@simova_enigma_2017], sites were excluded from analysis if they were managed, defined as plots that were planted, managed as plantations, irrigated, fertilised or including the term "managed" in their site description. 
Sites that had experienced significant disturbance within the past 100 years were also excluded. Disturbances that qualified sites for exclusion included major cutting or harvesting, burning, flooding, drought and storm events with site mortality >10% of trees. Grazed sites were retained.

@@

These findings clarify how FACF vary with latitude and climate on a global scale. Past studies have differed in their conclusions regarding the relationship between FACF and latitude or its correlates (Table 1, *H1*)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF are unambiguously highest in the tropical regions, and the relationship with latitude is approximately linear (Fig. 2). This contrasts with the suggestion that productivity of temperate forests is similar to that of tropical forests [@huston_global_2009]. Compared to tropical forests, the temperate forest biome has experienced more widespread anthropogenic disturbance and has a larger fraction of secondary stands [@potapov_mapping_2008; @poulter_global_2018], so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects. In addition, because carbon allocation varies with stand age [@delucia_forest_2007] (**Doughty et al. 2018: DOI: 10.1111/btp.12504; Anderson-Teixeira et al. 2013: doi: 10.1111/gcb.12194**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration (**Doughty et al. 2018: DOI: 10.1111/btp.12504**);[@delucia_forest_2007; @piao_forest_2010; @kunert_understanding_2019].

We show that FACF are broadly consistent in their responses to climate drivers on the global scale, with at most modest trends in C allocation among the variable pairs tested (Figs. 2, S3). 
This parallels the observation that C allocation across multiple FACF varies little with respect to climate along a steep tropical elevational gradient [@malhi_variation_2017; but see @moser_elevation_2011], 
and is not surprising given that trees face universal constraints in carbon allocation (**REFS**). 
The one trend in allocation that emerged from our analysis was a tendency for greater C allocation aboveground--and to foliage in particular--in warm tropical climates (Fig. S3). This is consistent with observations of increasing allocation to roots with declining temperature across a tropical elevational gradient [@moser_elevation_2011], 
and with observations and theory predicting relatively higher belowground carbon allocation at higher latitudes [@gill_belowground_2016].
It is also consistent with findings that as forest productivity increases, proportionaly more carbon is allocated to $ANPP_{foliage}$ relative to $BNPP_{fine.root}$ (*Chen et al. 2019; doi: 10.1111/ele.13193*) or $ANPP_{stem}$ [@hofhansl_new_2015]. **( check / comment on  @litton_carbon_2007 )**

One previously hypothesized trend that was not supported by our analysis was that tropical forests tend to have low carbon use efficiency ($CUE$= $NPP$/$GPP$=($GPP$-$R_{auto}$/$GPP$), which is based on observations of low $CUE$ in old-growth tropical forests relative to (mostly younger) extratropical forests [@delucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016]. Our analysis, limited to mature forests, shows no such trend (Fig. S3). $CUE$ is known to decline with forest age [@delucia_forest_2007; @piao_forest_2010; @collalti_is_2019], but appears to be roughly independent of $GPP$ [@litton_carbon_2007]. (*This finding may have some important implications for modeling and our broader understanding. I'm not sure offhand how much theory has been built on the idea of low CUE in tropical forests because of warm temperatures...Helene is not aware of any.*)

**update this paragraph ([issue 78](https://github.com/forc-db/Global_Productivity/issues/78))** 
One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$) than in subsidiary fluxes ($BNPP_{fine.root}$, $R_{root}$, $ANPP_{stem}$) as quantified by R^2^ (Fig. 2; Table S2). There are two, non-exclusive, potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may use fundamentally different approaches in different sites (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{fine.root}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled (*Clark et al. 2001 Ecological Applications*). However, methodological variation and uncertainty affect all of fluxes considered here, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more direct climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and respiration and directly through any climatic influence on C allocation, as well as many other local- and regional-scale factors (**REFS**).

The latitudinal gradient in FACF (Fig. 2) is driven primarily by temperature-related climate variables, the effects of which are moderated by moisture availability (Table 1, *H2-H3*; Figs. 3-4).  MAT and MAP have long been identified as primary global-scale drivers of FACF [*Lieth 1973: doi 10.1007/BF01536729*; **REFS**; @taylor_temperature_2017]. It is not appropriate to attempt to identify individual mean annual climate variables as mechanistic drivers of FACF because many climate variables co-vary across the latitudinal gradient (Fig. S2), because climatic drivers affect forest carbon flux on much shorter time scales than can be captured by annual climate summary variables, and because both climatic conditions and C flux vary intra- and inter-annually around the long-term means. However, it remains informative to consider these relationships. Among the temperature-related climate variables, $MAT$ is generally the most strongly correlated with FACF (Table S2)--perhaps in part because site-specific MAT is recorded for the majority of sites in ForC, whereas other variables were extracted from global gridded data products, introducing spatial scaling errors.

@@

FACF are negatively correlated with temperature seasonality (Table 1, *H4*; Fig. 4e), and are minimal during cold- or dry- dormant seasons. To account for this, a number of analyses seeking to characterize global-scale effects of climate on productivity have examined the relationship of C flux per month of the growing season with growing season climatic conditions [Table 1, *H5*; @kerkhoff_plant_2005; *Anderson et al. 2006: doi:10.1111/j.1461-0248.2006.00914.x *; *Enquist et al. 2007: doi: 10.1111/j.1365-2486.2006.01222.x*;  @michaletz_convergence_2014]. We found that the sort of simple metric needed to define growing season at a global scale [@kerkhoff_plant_2005] was rough **BBL: ?** with respect to temperature and problematic with respect to moisture because it doesn't capture temporal lags between precipitation and plant water use caused by storage in soil or snow. A temperature-defined growing season length had stong positive correlation with FACF (Fig. 4f), but explained less variation than $MAT$. Dividing FACFs by growing season length to yield FACF per growing season month removed the majority of climate-related variation, supporting the idea that the latitudinal gradient in FACF is attributable more to shorter growing seasons at high latitudes than to inherently lower rates of photosynthesis or respiration by high-latitude forests (*Enquist et al. 2007: doi: 10.1111/j.1365-2486.2006.01222.x*). However, there remained a number of significant correlations with growing season climatic conditions, suggesting that climatic conditions remain influential within the growing season. We conclude that while correcting for growing season length takes analyses a step closer to mechanistic linkage of instantaneous C flux rates to environmental conditions, it remains crude relative to the timescales on which climate affects plant metabolism, and does not advance statistical predictive power. Mechanistic accounting for climatic effects on global FACF patterns instead requires models representing physiologically meaningful timescales (*e.g.*, **refs**). **BBL: maybe cite e.g. ED2? http://dx.doi.org/10.5194/gmd-12-4309-2019 **

Our analysis clarifies how FACF vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of FACF (Figs. 2-3). This is consistent with observations of continental- to global-scale increases over time in $GPP$ [@li_mapping_2019] and $ANPP_{stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. However, increasing C flux rates are by no means universal [@rutishauser_testing_2020] (**MORE REFS**), likely because other factors are at play, including changes to other aspects of climate, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. **(discuss new Sullivan paper that finds higher max temperatures associated to lower ANPP_stem)**
Moreover, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapated and acclimatized will not necessarily parallel responses across geographic gradients in climate. Nevertheless, as we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.

>>>>

Carbon cycling in forests worldwide provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity and many human populations [@millennium_ecosystem_assessment_ecosystems_2005], while strongly influencing atmospheric CO~2~ and climate [@bonan_forests_2008]. Forests' autotrophic carbon fluxes (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance [@zak_plant_1994; @niedzialkowska_species_2010] and possibly diversity [@waide_relationship_1999; @chu_direct_2018]. FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients.

@@

The latitudinal gradient in FACF, along with altitudinal gradients [@girardin_net_2010; @malhi_variation_2017], is primarily driven by climate, which is a significant driver of FACF across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011; @hursh_sensitivity_2017]. The majority of studies have focused on exploring the relationships of FACF to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. There is strong evidence that both MAT and MAP show significant positive relationships with FACF [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of FACF across studies. Various measures of primary productivity {**FACF?**} saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity {**FACF?**} are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while $GPP$ increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] finds that $NPP$ increases linearly with temperature. Taylor et al. [-@taylor_temperature_2017] showed a positive interaction between MAT and MAP in shaping tropical forest productivity, such that high rainfall had a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates. Such complicated dynamics play out with belowground fluxes as well, with typically dominant soil temperature mediated by moisture and carbon supply [@hursh_sensitivity_2017; @xu_contribution_2016].

@@

This analysis focused on mature forests with no known history of signficant disturbance or management. 
There is evidence that stand age influences patterns of FACF and carbon allocation in forest ecosystems, and can confound relationships between latitude and primary productivity [@de_lucia_forest_2007; @gillman_latitude_2015]. To reduce any biasing effects of stand age, we included only stands of known age $\ge$ 100 years and those described by terms such as "mature", "intact", or "old-growth". 
Since management can alter observed patterns of FACF [@simova_enigma_2017], sites were excluded from analysis if they were managed, defined as plots that were planted, managed as plantations, irrigated, fertilised or including the term "managed" in their site description. 
Sites that had experienced significant disturbance within the past 100 years were also excluded. Disturbances that qualified sites for exclusion included major cutting or harvesting, burning, flooding, drought and storm events with site mortality >10% of trees. Grazed sites were retained.

@@

These findings clarify how FACF vary with latitude and climate on a global scale. Past studies have differed in their conclusions regarding the relationship between FACF and latitude or its correlates (Table 1, *H1*)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF are unambiguously highest in the tropical regions, and the relationship with latitude is approximately linear (Fig. 2). This contrasts with the suggestion that productivity of temperate forests is similar to that of tropical forests [@huston_global_2009]. Compared to tropical forests, the temperate forest biome has experienced more widespread anthropogenic disturbance and has a larger fraction of secondary stands [@potapov_mapping_2008; @poulter_global_2018], so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects. In addition, because carbon allocation varies with stand age [@de_lucia_forest_2007; @anderson-teixeira_altered_2013; @doughty_what_2018], age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration [@de_lucia_forest_2007; @piao_forest_2010; @doughty_what_2018; @kunert_understanding_2019].

We show that FACF are broadly consistent in their responses to climate drivers on the global scale, with at most modest trends in C allocation among the variable pairs tested (Figs. 2, S3). 
This parallels the observation that C allocation across multiple FACF varies little with respect to climate along a steep tropical elevational gradient [@malhi_variation_2017; but see @moser_elevation_2011], 
and is not surprising given that trees face universal constraints in carbon allocation (**REFS**). 
The one trend in allocation that emerged from our analysis was a tendency for greater C allocation aboveground--and to foliage in particular--in warm tropical climates (Fig. S3). This is consistent with observations of increasing allocation to roots with declining temperature across a tropical elevational gradient [@moser_elevation_2011], 
and with observations and theory predicting relatively higher belowground carbon allocation at higher latitudes [@gill_belowground_2016].
It is also consistent with findings that as forest productivity increases, proportionaly more carbon is allocated to $ANPP_{foliage}$ relative to $BNPP_{fine.root}$ [@chen_allometry_2019] or $ANPP_{stem}$ [@hofhansl_new_2015]. **( check / comment on  @litton_carbon_2007 )**

One previously hypothesized trend that was not supported by our analysis was that tropical forests tend to have low carbon use efficiency ($CUE$= $NPP$/$GPP$=($GPP$-$R_{auto}$/$GPP$), which is based on observations of low $CUE$ in old-growth tropical forests relative to (mostly younger) extratropical forests [@de_lucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016]. Our analysis, limited to mature forests, shows no such trend (Fig. S3). $CUE$ is known to decline with forest age [@de_lucia_forest_2007; @piao_forest_2010; @collalti_is_2019], but appears to be roughly independent of $GPP$ [@litton_carbon_2007]. (*This finding may have some important implications for modeling and our broader understanding. I'm not sure offhand how much theory has been built on the idea of low CUE in tropical forests because of warm temperatures...Helene is not aware of any.*)

**update this paragraph ([issue 78](https://github.com/forc-db/Global_Productivity/issues/78))** 
One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$) than in subsidiary fluxes ($BNPP_{fine.root}$, $R_{root}$, $ANPP_{stem}$) as quantified by R^2^ (Fig. 2; Table S2). There are two, non-exclusive, potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may use fundamentally different approaches in different sites (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{fine.root}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled [@clark_measuring_2001]. However, methodological variation and uncertainty affect all of fluxes considered here, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more direct climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and respiration and directly through any climatic influence on C allocation, as well as many other local- and regional-scale factors (**REFS**).

The latitudinal gradient in FACF (Fig. 2) is driven primarily by temperature-related climate variables, the effects of which are moderated by moisture availability (Table 1, *H2-H3*; Figs. 3-4).  MAT and MAP have long been identified as primary global-scale drivers of FACF [@lieth_primary_1973; **REFS**; @taylor_temperature_2017]. It is not appropriate to attempt to identify individual mean annual climate variables as mechanistic drivers of FACF because many climate variables co-vary across the latitudinal gradient (Fig. S2), because climatic drivers affect forest carbon flux on much shorter time scales than can be captured by annual climate summary variables, and because both climatic conditions and C flux vary intra- and inter-annually around the long-term means. However, it remains informative to consider these relationships. Among the temperature-related climate variables, $MAT$ is generally the most strongly correlated with FACF (Table S2)--perhaps in part because site-specific MAT is recorded for the majority of sites in ForC, whereas other variables were extracted from global gridded data products, introducing spatial scaling errors.

@@

FACF are negatively correlated with temperature seasonality (Table 1, *H4*; Fig. 4e), and are minimal during cold- or dry- dormant seasons. To account for this, a number of analyses seeking to characterize global-scale effects of climate on productivity have examined the relationship of C flux per month of the growing season with growing season climatic conditions [Table 1, *H5*; @kerkhoff_plant_2005; @anderson_temperature-dependence_2006; @enquist_adaptive_2007; @michaletz_convergence_2014]. We found that the sort of simple metric needed to define growing season at a global scale [@kerkhoff_plant_2005] was rough **BBL: ?** with respect to temperature and problematic with respect to moisture because it doesn't capture temporal lags between precipitation and plant water use caused by storage in soil or snow. A temperature-defined growing season length had stong positive correlation with FACF (Fig. 4f), but explained less variation than $MAT$. Dividing FACFs by growing season length to yield FACF per growing season month removed the majority of climate-related variation, supporting the idea that the latitudinal gradient in FACF is attributable more to shorter growing seasons at high latitudes than to inherently lower rates of photosynthesis or respiration by high-latitude forests [@enquist_adaptive_2007]. However, there remained a number of significant correlations with growing season climatic conditions, suggesting that climatic conditions remain influential within the growing season. We conclude that while correcting for growing season length takes analyses a step closer to mechanistic linkage of instantaneous C flux rates to environmental conditions, it remains crude relative to the timescales on which climate affects plant metabolism, and does not advance statistical predictive power. Mechanistic accounting for climatic effects on global FACF patterns instead requires models representing physiologically meaningful timescales (*e.g.*, **refs**) [@longo_biophysics_2019]

Our analysis clarifies how FACF vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of FACF (Figs. 2-3). This is consistent with observations of continental- to global-scale increases over time in $GPP$ [@li_mapping_2019] and $ANPP_{stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. **DISCUSS/CITE [Wu et al. 2020]** [@wu_accelerated_2020] However, increasing C flux rates are by no means universal [@rutishauser_testing_2020] (**MORE REFS**), likely because other factors are at play, including changes to other aspects of climate, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. **(discuss new Sullivan paper that finds higher max temperatures associated to lower ANPP_stem)**
Moreover, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapated and acclimatized will not necessarily parallel responses across geographic gradients in climate. Nevertheless, as we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.
